Synthesis of Rules for Network Intrusion Detectors

### 2019-10-24 paper/paper.tex
another pass

\keywords{NIDS, synthesis, search}
\end{abstract}
%
%
%
\section{Introduction}

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial NIDS implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based and anomaly-based \nids. A
rule-based intrusion detector\footnote{\aka\ signature-based intrusion
  detector.} checks if the network traffic matches a fixed set of
rules. Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{}. This rule
prescribes a method to capture a denial-of-service
attack~\cite{understanding-dos} to a server by matching specific
conditions about the network traffic. Relevant properties about the
traffic of interest appear in bold (see
Section~\ref{sec:suri-metas-coverage}). Rule-based \nids{} focus on
known attacks whereas anomaly-based \nids{} focus on unknown attacks
that could be observed by first learning the regular traffic and then
acting
properly~\cite{kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. These
two kinds of intrusion detection mechanisms are
complementary. Rule-based NIDS could miss attacks (as rulesets can
become outdated) whereas anomaly-based NIDS can raise false alarms (as
detection algorithms are approximate).

Rule-based NIDS are restricted to a fixed set of rules defined by the
network system administrator. Manually writing these rules is tedious,
error-prone, but important as attackers constantly create new
strategies to circumvent existing rules. IT-security companies
capitalize on this phenomena and put rulesets on the market for
sale~\cite{proofpoint-etpro,snort-rule-subscriptions}.  Despite this
fundamental issue, rule-based systems are tremendously popular in
industry today. Anomaly-based NIDS are not restricted to a fixed set
of rules. They learn usage behavior from positive traffic and, based
on that, alert uncommon behavior~\cite{7579764}. Unfortunately,
anomaly-based NIDS are fundamentally imprecise. To sum up, rule-based
NIDS are precise but incomplete whereas anomaly-based NIDS are (more)
complete but imprecise.

%% This paper focuses on the
%% problem of discovering new rules by using anomaly-based NIDS.

This paper proposes \tname{}, a technique that uses machine
intelligence to automatically synthesize rules for rule-based
\nids. Our general goal is to harden the protection of NIDS by
improving the creation process of rules. One scenario of application
of \tname{} is one where an anomaly-based NIDS identifies abnormal
traffic and signals that traffic to \tname{} to create and distribute
the rules.

%% Note that anomaly-based NIDS are evolving pretty quick with the
%% advances in machine learning, but rule-based NIDS are still extremely
%% popular.  \tname{} could also leverage existing databases of malicious
%% traffic to synthesize rules.  Regardless of how the negative traffic
%% is produced (out of scope of this paper), those rules can be
%% distributed to rule databases for free.
%%  The
%% core motivation is that 1) attackers are productive in creating new
%% ways to circumvent existing protections and 2) manual creation of
%% rules is tedious and time-consuming.

@@

Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.
\begin{wrapfigure}[15]{r}{0.5\textwidth}
  \centering
  \vspace{-5ex}
  \scalebox{0.7}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-1ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
\end{wrapfigure}
\MyComment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, column ``\#~Params'' shows the number of
additional parameters associated with a given option, and column
``Description'' presents a short description of the option purpose. We
found that the option \CodeIn{content} is particularly prevalent in
rules from public databases (see
Figure~\ref{fig:distribution-contents}). This option takes a sequence
of bytes as argument (\eg{}, \CodeIn{content:''SELECT ''}) and checks
if that sequence is present in the payload of the message. Detailed
description of the Suricata rules can be found
elsewhere~\cite{suri-rule-format}.

\begin{table}[h]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{p{0.35cm}ll}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

@@

\Fix{Illustrate how information is extracted from packets}

@@

This section briefly presents examples of attacks and rules that are
synthesized with \tname{} to capture those attacks.

@@

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the \Fix{...what source...}
The relevant options in the rule appear in bold---the option
\CodeIn{flags: S,12}, identifies a SYN packet in a TCP packet and the
option \CodeIn{threshold: type both, track by\_dst, count 5000,
  seconds 5} indicates that a high volume of such packets should be
requested in a short period of time.

@@

\tname{} is a technique that uses positive and negative examples to
synthesize signature-based NIDS rules, such as those from Snort and
Suricata. The goal of \tname\ is to create a rule that \emph{only}
captures the positive traffic, \ie{}, the malicious sequence of
messages.

@@

\subsection{Step 1: Reverse Engineering}
\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}

\begin{wrapfigure}[14]{r}{0.56\textwidth}
  \centering
  \vspace{-8ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
The goal of the first step is to extract options from the positive
traffic. For that, \tname{} parses the protocol message, looking for
fields in the message associated with options (see
Section~\ref{sec:rules-and-packets}). For example, fields
\Fix{...and...}  can be used to determine the option
\Fix{zzz}. Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content}
options. Figure~\ref{fig:distribution-contents} shows the histogram of
number of these options per rule (up to 10) for the Suricata public
rule database \Fix{cite}. Rules without \CodeIn{content} options is an
exception and half of the rules contain at least two \CodeIn{content}
options.  To find the values associated with these options, \tname{}
splits the payload of the malicious message in tokens using natural
language delimiters, such as $\backslash$t, $\backslash$n,
$\backslash$r, and spaces. The rationale for this decision is that we
observed, by inspecting existing rules, that \CodeIn{content} options
typically refer to sequence of characters in the payload of messages
separated by these delimiters. This tokenization step can generate
dozens of options. For the example from
Section~\ref{sec:content-example}, \tname{} produces a total of
\Fix{xx} content options.

>>>>

\keywords{\nids, synthesis, search}
\end{abstract}
%
%
%
\section{Introduction}

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial \nids implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based and anomaly-based \nids. A
rule-based intrusion detector\footnote{\aka\ signature-based intrusion
  detector.} checks if the network traffic matches a fixed set of
rules. Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{}. This rule
prescribes a method to capture a denial-of-service
attack~\cite{understanding-dos} to a server by matching specific
conditions about the network traffic. Relevant properties about the
traffic of interest appear in bold (see
Section~\ref{sec:suri-metas-coverage}). Rule-based \nids{} focus on
known attacks whereas anomaly-based \nids{} focus on unknown attacks
that could be observed by first learning the regular traffic and then
acting
properly~\cite{kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. These
two kinds of intrusion detection mechanisms are
complementary. Rule-based \nids could miss attacks (as rulesets can
become outdated) whereas anomaly-based \nids can raise false alarms (as
detection algorithms are approximate).

Rule-based \nids are restricted to a fixed set of rules defined by the
network system administrator. Manually writing these rules is tedious,
error-prone, but important as attackers constantly create new
strategies to circumvent existing rules. IT-security companies
capitalize on this phenomena and put rulesets on the market for
sale~\cite{proofpoint-etpro,snort-rule-subscriptions}.  Despite this
fundamental issue, rule-based systems are tremendously popular in
industry today. Anomaly-based \nids are not restricted to a fixed set
of rules. They learn usage behavior from positive traffic and, based
on that, alert uncommon behavior~\cite{7579764}. Unfortunately,
anomaly-based \nids are fundamentally imprecise. To sum up, rule-based
\nids are precise but incomplete whereas anomaly-based \nids are (more)
complete but imprecise.

%% This paper focuses on the
%% problem of discovering new rules by using anomaly-based \nids.

This paper proposes \tname{}, a technique that uses machine
intelligence to automatically synthesize rules for rule-based
\nids. Our general goal is to harden the protection of \nids by
improving the creation process of rules. One scenario of application
of \tname{} is one where an anomaly-based \nids identifies abnormal
traffic and signals that traffic to \tname{} to create and distribute
the rules.

%% Note that anomaly-based \nids are evolving pretty quick with the
%% advances in machine learning, but rule-based \nids are still extremely
%% popular.  \tname{} could also leverage existing databases of malicious
%% traffic to synthesize rules.  Regardless of how the negative traffic
%% is produced (out of scope of this paper), those rules can be
%% distributed to rule databases for free.
%%  The
%% core motivation is that 1) attackers are productive in creating new
%% ways to circumvent existing protections and 2) manual creation of
%% rules is tedious and time-consuming.

@@

\begin{wrapfigure}[14]{r}{0.5\textwidth}
  \centering
  \vspace{-7ex}
  \scalebox{0.7}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-2ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
\end{wrapfigure}
Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.\MyComment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column
``Description'' presents a short description of the option purpose. We
found that the option \CodeIn{content} is particularly prevalent in
rules from public databases (see
Figure~\ref{fig:distribution-contents}). This option takes a sequence
of bytes as argument (\eg{}, \CodeIn{content:''SELECT ''}) and checks
if that sequence is present in the payload of the message. Detailed
description of the Suricata rules can be found
elsewhere~\cite{suri-rule-format}.

\begin{table}[t!]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{p{0.35cm}ll}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{figure*}[h!]
\centering
\includegraphics[scale=0.3]{figs/ICMP-packet-structure.png}
\caption{Layout of an ICMP packet.}
\label{fig:icmp-packet-layout}
\end{figure*}

There exists a correspondence between network packets and most rule
options. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. (Table~\ref{table:rules} lists some
example rule options.) The values of the options ``ttl'', ``itype'',
and ``icode'' are the same as the value of the corresponding fields in
the ICMP packet. For the field ``Flags'' the value of the
corresponding option is encoded in its bitvector. For example, if the
``Flags'' field stores the value 1010, the corresponding rule option
will be \CodeIn{flagbits:RM}, as each position of the vector encodes a
different character in the option \CodeIn{flagbits}.

The value of most options can be obtained through transformation
functions, as those described above. However, this is not the case for
the option \CodeIn{threshold}, which is typically used in
denial-of-service attacks. Conceptually, that option operates over
sequence of messages as opposed to a single message. More precisely,
to trigger an action on a rule containing that option the \nids\ tool
needs to monitor several messages.

%% \Luc{
%%   Os campos Flags e Time to live no header IP são mapeados para as opções fragbits e ttl, respectivamente. Os campos Type e Code no header ICMP são mapeados para as opções itype e icode. \\
%% Os campos, Time to Live, Type e Code recebem diretamente o valor que está no pacote. Por exemplo, se o campo Type no pacote possuir valor 8, na regra ele vai aparecer como itype:8.\\
%% O campo Flags é intepretado como um vetor de bits, onde cada posição representa uma flag. Na regra, o valor do campo Flags é traduzido em caracteres que representam cada flag presente no pacote. Por exemplo: se o campo Flags possuir o valor 1010, na regra aparecerá fragbits:RM, pois os caracteres R e M indicam a presença do quarto e segundo bits.}

@@

This section briefly presents examples of attacks and illustrates the
rules that \tname{} synthesizes to capture those attacks.

@@

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a TCP packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of time.

@@

\tname{} is a technique that uses positive and negative examples to
synthesize signature-based \nids rules, such as those from Snort and
Suricata. The goal of \tname\ is to create a rule that \emph{only}
captures the positive traffic, \ie{}, the malicious sequence of
messages.

@@

\subsection{Step 1: Reverse Engineering}
\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}
The goal of the first step is to extract options from the positive
traffic. For that, \tname{} parses the protocol message, looking for
fields in the message associated with options (see
Section~\ref{sec:rules-and-packets}).
Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content}
options. Figure~\ref{fig:distribution-contents} shows the histogram of
number of these options per rule (up to 10) obtained from a public
ruleset~\cite{emerging-threats-open} containing 27.8K Suricata
rules. Rules without \CodeIn{content} options is an exception and half
of the rules contain at least two \CodeIn{content} options.  To find
\begin{wrapfigure}[15]{r}{0.56\textwidth}
  \centering
  \vspace{-4ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
the values associated with these options, \tname{} splits the payload
of the malicious message in tokens using natural language delimiters,
such as $\backslash$t, $\backslash$n, $\backslash$r, and spaces. The
rationale for this decision is that we observed, by inspecting
existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. This tokenization step can generate dozens of options. For
the example from Section~\ref{sec:content-example}, \tname{} produces
a total of \Fix{xx} content options.
